\textcolor[rgb]{0,0,1}{Data-Driven Model Predictive Control for Energy and Thermal Comfort Optimization in Building Energy Systems using Random Forests}

### 2018-01-19 paper/final/cs-dpc.tex
Updated Sec2,3,5, and intro review by Ale

In this section, we explain how DPC can be applied to this case study. We begin with a description of features $\tX$ and output $\tY$ used for training.

@@

The fundamental reason why DPC  is suitable for such a problem is that when the complexity rises, there is a huge cost to model all the states given by the dynamical system \eqref{E:bilinear1}. For example, states in the bilinear model also include slab temperatures which require modeling of structural and material properties in detail and often we also need to install new sensors to capture additional states. Thus, DPC is based solely on one state of the model, i.e. the zone temperature that can be easily measured with a thermostat. This serves as the output variable $\tY$ of interest for which we build $N$ trees and $N$ forests as described in Section~\ref{SS:dpcrt} and \ref{SS:dpcrf}, respectively. Therefore, $\tY_{\mathrm{k+j|k}}:=x_{\mathrm{k+j|k}}^1$, where $x^1$ is the first component of $x$.
Next, we define the non-manipulated features $\tX^d_{\mathrm{k|k}}$. At time $k$, for the tree $\mathcal{T}_j$ and the forest $\mathcal{R}_j$, we base these features to include weather disturbances, external disturbances due to occupancy and equipments, and autoregressive terms of the room temperature, i.e.
$\tX^d_{\mathrm{k|k}}:=[ d_{\mathrm{k+j-N|k}} ,\dots,d_{\mathrm{k+j-1|k}}, x_{\mathrm{k|k}}^1,\dots, x_{\mathrm{k-\delta|k}}^1 ]$, where $\delta$ is the order of autoregression.
Finally, the inputs in DPC are exactly same as in MPC. i.e. $\tX^c_{\mathrm{k+j-1|k}}:=u_{\mathrm{k+j-1|k}}$.
The training data in the above format was generated by simulating the bilinear model with rule-based strategies for 10 months in 2007. January and May were deliberately excluded for testing the DPC implementation.
\subsubsection{Optimization} 
\label{SSS:dpc_opt}
For a fair comparison with MPC, we cast DPC optimization problem as follows:

\begin{subequations}
\begin{align}
\text{min } & \sum_{j=1}^{N} ({\tY}_{\mathrm{k+j|k}} - x_{ref}) \mathcal{Q}^{(1,1)} ({\tY}_{\mathrm{k+j|k}} - x_{ref}) + c^T \tX^c_{\mathrm{k+j-1|k}}+  \lambda\epsilon_j\\
\text{s.~t. } & \ \ \ \ \ \tY_{\mathrm{k+j|k}} =  \alpha_j^T \left[1,\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+j-1|k}} \right]^T \label{SE:dpc}\\
& \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \underline{\tX}^c \leq \tX^c_{\mathrm{k+j-1|k}} \leq \bar{\tX}^c\\ 
& \ \ \ \ \ \ \ \ \ \ \ \ \underline{\tY}-\epsilon_j \leq \tY_{\mathrm{k+j|k}} \leq \bar{\tY} + \epsilon_j\\\
& \ \ \ \ \ \ \ \ \ \ \ \ \ \ \epsilon_j \geq 0, \ j = 1,\dots,N.
\end{align} \label{E:dpc}
\end{subequations}
\noindent Here $\alpha = \beta$ for DPC-RT and $\alpha = \hat{\Theta}$ for DPC-En.
Note that, \eqref{E:dpc} is DPC analog of \eqref{E:mpc}. The only difference is the state dynamics \eqref{SE:mpc1} and \eqref{SE:mpc2} are now replaced with \eqref{SE:dpc}.

@@

We compare the prediction for the first time step $\tY_{\mathrm{k+1|k}}$ and the 6-hour ahead prediction $\tY_{\mathrm{k+6|k}}$ for a week in the month of May in Figure~\ref{F:validation}. It is visible how trees have a high variance, and the forests are more accurate. Note that data from January and May were not used for training. The quantitative summary of the accuracy is given in Table~\ref{T:validation}. We can see that the random forests are better in all respects.
\begin{table}[h!]
	\centering
	\begin{tabular}{cccc}
		\toprule
		& RMSE & $R^2$ score & EV  \\ 
		\midrule
		tree-$\tY_{\mathrm{k+1|k}}$    &  0.42 &  0.75 & 0.76    \\
		tree-$\tY_{\mathrm{k+6|k}}$  & 0.64 &  0.41  & 0.42 \\
		forest-$\tY_{\mathrm{k+1|k}}$  & 0.29 & 0.87  & 0.88 \\
		forest-$\tY_{\mathrm{k+6|k}}$  & 0.38 & 0.78 & 0.80 \\
		\bottomrule
	\end{tabular}
	\caption{Quantitative comparison of root mean square error (RMSE), $R^2$ score, and explained variance (EV) for trees and forests for different predictions steps.}
	\captionsetup{justification=centering}
	\label{T:validation}
\end{table}

>>>>

In this section, we explain how DPC can be applied to this case study. We begin with a description of features $\X$ and output $\Y$ used for training.

@@

The fundamental reason why DPC  is suitable for such a problem is that when the complexity rises, there is a huge cost to model all the states given by the dynamical system \eqref{E:bilinear1}. For example, states in the bilinear model also include slab temperatures which require modeling of structural and material properties in detail and often we also need to install new sensors to capture additional states. Thus, DPC is based solely on one state of the model, i.e. the zone temperature that can be easily measured with a thermostat. This serves as the output variable in $\Y$ of interest for which we build $N$ trees and $N$ forests as described in Section~\ref{SS:dpcrt} and \ref{SS:dpcrf}, respectively. Therefore, $y_{k+j}:=x_{k+j+1}^1$, where $x^1$ is the first component of $x$.
Next, we define the non-manipulated features in $\X^d$. At time $k$, for the tree $\mathcal{T}_j$ and the forest $\mathcal{F}_j$, we base these features to include weather disturbances, external disturbances due to occupancy and equipments, and autoregressive terms of the room temperature, i.e.
$\X^d:=\{( d_{k+j-N} ,\ldots,d_{k+j-1}, x_{k}^1,\ldots, x_{k-\delta}^1 )\},\ \forall k=1,\ldots,n$, where $\delta$ is the order of state autoregression.
Finally, the inputs in DPC are exactly same as in MPC, i.e. $\X^c:=u_k,\ \forall k=1,\ldots,n$.
The training data in the above format was generated by simulating the bilinear model with rule-based strategies for 10 months in 2007. January and May were deliberately excluded for testing the DPC implementation.
\subsubsection{Optimization} 
\label{SSS:dpc_opt}
For a fair comparison with MPC, we cast DPC optimization problem as follows:

\begin{subequations}
	\begin{align}
	& \underset{u_{k+j},\epsilon_j}{\text{minimize}} & & \sum_{j=1}^{N} (y_{k+j} - x_{ref}) \mathcal{Q}^{(1,1)} (y_{k+j} - x_{ref}) + c^T u_{k+j-1} + \lambda\epsilon_j \\
	& \text{subject to }                  & & y_{k+j} =  \alpha_j \left[1,u_{k},\ldots,u_{k+j-1} \right]^\top \label{SE:dpc}                                            \\
	&                                     & & \underline{u} \leq u_{k+j-1} \leq \bar{u} 									   	                                        \\
	&                                     & & \underline{y} - \epsilon_j \leq y_{k+j} \leq \bar{y} + \epsilon_j                                                         \\
	&                                     & & \epsilon_j \geq 0, \ j = 1,\dots,N,            									                 
	\end{align}\label{E:dpc}
\end{subequations}

\noindent where $\alpha = \beta$ for DPC-RT and $\alpha = \hat{\Theta}$ for DPC-En.
Note that, \eqref{E:dpc} is DPC analog of \eqref{E:mpc}. The only difference is the state dynamics \eqref{SE:mpc1} and \eqref{SE:mpc2} are now replaced with \eqref{SE:dpc}.

@@

We compare the prediction for the first time step $y_{k+1}$ and the 6-hour ahead prediction $y_{k+6}$, given $k$, for a week in the month of May in Figure~\ref{F:validation}. It is visible how trees have a high variance, and the forests are more accurate. Note that data from January and May were not used for training. The quantitative summary of the accuracy is given in Table~\ref{T:validation}. We can see that the random forests are better in all respects.
\begin{table}[h!]
	\centering
	\begin{tabular}{cccc}
		\toprule
		                  & RMSE & $R^2$ score & EV  \\ 
		\midrule
		tree-$y_{k+1}$    & 0.42 & 0.75        & 0.76 \\
		tree-$y_{k+6}$    & 0.64 & 0.41        & 0.42 \\
		forest-$y_{k+1}$  & 0.29 & 0.87        & 0.88 \\
		forest-$y_{k+6}$  & 0.38 & 0.78        & 0.80 \\
		\bottomrule
	\end{tabular}
	\caption{Quantitative comparison of root mean square error (RMSE), $R^2$ score, and explained variance (EV) for trees and forests for different predictions steps.}
	\captionsetup{justification=centering}
	\label{T:validation}
\end{table}

### 2018-01-19 paper/final/cs-model.tex
Updated Sec2,3,5, and intro review by Ale

The model dynamics are given below. The bilinearity is present in both input-state, and input-disturbance.
\begin{gather}
\label{E:bilinear1}
x_{k+1} = Ax_{k}+(B_u +B_{xu}[x_k] + B_{du}[d_k]) u_k+B_dd_k \\
x_{k} \in \mathbb{R}^{12}, u_{k} \in \mathbb{R}^{4}, d_{k} \in \mathbb{R}^{8} \ \forall k = 0,\dots,T, \nonumber
\end{gather}
where, the matrices $B_{xu}$ and $B_{du}$ are defined as
\begin{gather}
\label{E:bilinear2}
B_{xu}[x_k] = [ B_{xu,1}[x_k],B_{xu,2}[x_k], \dots, B_{xu,4}[x_k] ] \in \mathbb{R}^{12\times4}, \nonumber\\
B_{du}[d_k] = [ B_{du,1}[d_k],B_{du,2}[d_k], \dots, B_{du,4}[d_k] ] \in \mathbb{R}^{12\times4}\nonumber, \\
B_{xu,i} \in \R^{12\times12}, B_{du,i} \in \R^{12\times8} \ \forall i=1,2,3,4.\nonumber
\end{gather}
For this study, we assume that the disturbances are precisely known to MPC as well as DPC controller. In our future work, we will account for the uncertainties in the disturbances with an extension to Scenario approach \cite{Bernardini2009} for DPC.

>>>>

The model dynamics are given below. The bilinearity is present in both input-state, and input-disturbance.
\begin{equation}\label{E:bilinear1}
x_{k+1} = Ax_{k}+(B_u +B_{xu}[x_k] + B_{du}[d_k]) u_k+B_dd_k,
\end{equation}
where $x_{k} \in \mathbb{R}^{12}, u_{k} \in \mathbb{R}^{4}, d_{k} \in \mathbb{R}^{8} \ \forall k = 0,\dots,T$, and the matrices $B_{xu}$ and $B_{du}$ are defined as
\begin{equation}
	\begin{aligned}
		&B_{xu}[x_k] = [ B_{xu,1}[x_k],B_{xu,2}[x_k], \dots, B_{xu,4}[x_k] ] \in \mathbb{R}^{12\times4}, \\
		&B_{du}[d_k] = [ B_{du,1}[d_k],B_{du,2}[d_k], \dots, B_{du,4}[d_k] ] \in \mathbb{R}^{12\times4}, \\
	\end{aligned}
\end{equation}
with $B_{xu,i} \in \R^{12\times12}, B_{du,i} \in \R^{12\times8} \ \forall i=1,2,3,4$.
For this study, we assume that the disturbances are precisely known to MPC as well as DPC controller.

### 2018-01-19 paper/final/cs-mpc.tex
first change fot the revision
Updated Sec2,3,5, and intro review by Ale

We use an MPC controller with a quadratic and a linear cost for comparison.
The finite RHC approach involves optimizing a cost function subject to the dynamics of the system and the constraints, over a finite horizon of time \cite{Mayne2000}. After an optimal sequence of control inputs are computed, the first input is applied, then at the next step the optimization is solved again.

@@

The objective of the controller is to minimize the energy usage $c^Tu$ while maintaining a desired level of thermal comfort $x_{ref}$.
Therefore, at time step $k$, we solve a continuously linearized MPC problem to determine the optimal sequence of inputs $[u_{\mathrm{k|k}},\dots,u_{\mathrm{k+N-1|k}}]$:
\begin{subequations}
\begin{align}
\text{min } & \sum_{j=1}^{N} ({x}^T_{\mathrm{k+j|k}} - x_{ref}) \mathcal{Q} ({x}_{\mathrm{k+j|k}} - x_{ref}) + c^Tu_{\mathrm{k+j-1}} +  \lambda\epsilon_j\\
\text{s.~t. } & \ \ x_{\mathrm{k+j|k}} =  Ax_{\mathrm{k|k}} + B u_{\mathrm{k+j-1|k}} + B_d d_{\mathrm{k+j-1|k}} \label{SE:mpc1} \\
& \ \ \ \ \ B = B_u + B_{xu}[x_{\mathrm{k|k}}] + B_{du}[d_{\mathrm{k+j-1|k}}] \label{SE:mpc2}\\
& \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \underline{u} \leq u_{\mathrm{k+j-1|k}} \leq \bar{u}\\ 
& \ \ \ \ \ \ \ \ \ \ \ \ \underline{x}-\epsilon_j \leq x_{\mathrm{k+j|k}} \leq \bar{x} + \epsilon_j\\\
& \ \ \ \ \ \ \ \ \ \ \ \ \ \ \epsilon_j \geq 0, \ j = 1,\dots,N,
\end{align}\label{E:mpc}
\end{subequations} 
\noindent where $\mathcal{Q} \in \R^{12 \times 12}$ has all zeros except at $\mathcal{Q}^{(1,1)}$ corresponding to the zone temperature, $c \in \R^{4}$ is proportional to cost of using each actuator and $\lambda$ penalizes the slack variables.

>>>>

We use an MPC controller with a quadratic and a linear cost for comparison.
The finite RHC approach involves optimizing a cost function subject to the dynamics of the system and the constraints, over a finite horizon of time \cite{Mayne2000}. After an optimal sequence of control inputs is computed, the first input is applied, then at the next step the optimization is solved again.

@@

The objective of the controller is to minimize the energy usage $c^Tu$ while maintaining a desired level of thermal comfort $x_{ref}$ (or $\tT_\mathrm{ref}$).
Therefore, at time step $k$, we solve a continuously linearized MPC problem to determine the optimal sequence of inputs $[u_{\mathrm{k|k}},\dots,u_{\mathrm{k+N-1|k}}]$:

\begin{subequations}
	\begin{align}
		& \underset{u_{k+j},\epsilon_j}{\text{minimize}} & & \sum_{j=1}^{N} ({x}_{k+j} - x_{ref})\top Q ({x}_{k+j} - x_{ref}) + c^Tu_{k+j-1} +  \lambda\epsilon_j \\
		& \text{subject to }                             & & x_{k+j} = Ax_{k+j-1} + B u_{k+j-1} + B_d d_{k+j-1} \label{SE:mpc1}                                   \\
		&                                                & & B       = B_u + B_{xu}[x_{k}] + B_{du}[d_{k+j-1}] \label{SE:mpc2}                                    \\
		&                                                & & \underline{u} \leq u_{k+j-1} \leq \bar{u} 									   	                      \\
		&                                                & & \underline{x} - \epsilon_j \leq x_{k+j} \leq \bar{x} + \epsilon_j                                    \\
		&                                                & & \epsilon_j \geq 0, \ j = 1,\dots,N,            									                 
	\end{align}\label{E:mpc}
\end{subequations}

\noindent where $Q \in \R^{12 \times 12}$ has all zeros except at $Q^{(1,1)}$ corresponding to the zone temperature, $c \in \R^{4}$ is proportional to cost of using each actuator and $\lambda$ penalizes the slack variables.
